---
title: "Regresión lineal múltiple"
author: "Luis Ramón Munive Hernández"
date: "11/19/2021"
output:
  pdf_document:
    keep_tex: yes
---

```{r}
datos <- data.frame(Carbohydrate = c(33,
40,
37,
27,
30,
43,
34,
48,
30,
38,
50,
51,
30,
36,
41,
42,
46,
24,
35,
37),
Age = c(33,
47,
49,
35,
46,
52,
62,
23,
32,
42,
31,
61,
63,
40,
50,
64,
56,
61,
48,
28),
Weight = c(100,
92,
135,
144,
140,
101,
95,
101,
98,
105,
108,
85,
130,
127,
109,
107,
117,
100,
118,
102),
Protein = c(14,
15,
18,
12,
15,
15,
14,
17,
15,
14,
17,
19,
19,
20,
15,
16,
18,
13,
18,
14))
```

```{r Visualización de datos, fig.width=3, fig.height=3, fig.keep='h'}
library(ggplot2)
ggplot(data = datos, mapping = aes(x = Age, y = Carbohydrate)) +
  geom_point() +
  xlab(label = "Edad") +
  ylab(label = "Carbohidratos")

ggplot(data = datos, mapping = aes(x = Weight, y = Carbohydrate)) +
  geom_point() +
  xlab(label = "Peso") +
  ylab(label = "Carbohidratos")

ggplot(data = datos, mapping = aes(x = Protein, y = Carbohydrate)) +
  geom_point() +
  xlab(label = "Proteína") +
  ylab(label = "Carbohidratos")
```

```{r}
y <- datos$Carbohydrate
X <- as.matrix(cbind(rep(1, nrow(datos)), datos[,-1]))
```

```{r}
b <- solve(t(X) %*% X) %*% (t(X) %*% y)
b
```

$\boldsymbol{b} = \left(\boldsymbol{X}^T \boldsymbol{X}\right)^{-1} \boldsymbol{X}^T \boldsymbol{y} = [$ `r t(as.vector(b))` $]^T$

$\boldsymbol{y}^T \boldsymbol{y} =$ `r t(y) %*% y`

$N \overline{y}^2 =$ `r nrow(X) * mean(y)^2`

$\boldsymbol{b}^T \boldsymbol{X}^T \boldsymbol{y} =$ `r (t(b) %*% t(X)) %*% y`

Suma de cuadrados residual $=$ `r t(y) %*% y - (t(b) %*% t(X)) %*% y`

```{r}
N <- nrow(X)
p <- ncol(X)
sigma2.hat <- as.numeric(1 / (N - p) * t(y - X %*% b) %*% (y - X %*% b))
sigma2.hat
```

$\widehat{\sigma}^2 = \dfrac{1}{N - p} \left(\boldsymbol{y} - \boldsymbol{X} \boldsymbol{b} \right)^T \left(\boldsymbol{y} - \boldsymbol{X} \boldsymbol{b} \right) =$ `r sigma2.hat`

```{r}
V.b <- sigma2.hat * solve(t(X) %*% X)
V.b
se.b <- sqrt(diag(V.b))
se.b
```

```{r}
library(knitr)
library(kableExtra)
kable(x = data.frame(termino = c("Constante", "Edad", "Peso", "Proteína"),
                     bj = b,
                     se = se.b),
      format = "latex",
      digits = 5,
      row.names = FALSE,
      col.names = c("Término", "$b_j$", "Error estándar"),
      align = c("l", "r", "r"), booktabs = TRUE, escape = FALSE) %>%
  kable_styling(position = "center", latex_options = "hold_position")
```

```{r}
X0 <- X[,-2]
b0 <- solve(t(X0) %*% X0) %*% (t(X0) %*% y)
b0
```

$\boldsymbol{b}_0 =[$ `r b0` $]^T$

$\boldsymbol{b}_0^T \boldsymbol{X}_0^T \boldsymbol{y} =$ `r (t(b0) %*% t(X0)) %*% y`

Suma de cuadrados residual $=$ `r t(y) %*% y - (t(b0) %*% t(X0)) %*% y`

```{r F value, fig.width=6, fig.height=4}
library(sjPlot)
dist_f(f = 38.359 / 35.47894, deg.f1 = 1, deg.f2 = 16)
```

```{r}
y.hat <- X %*% b
residual <- y - y.hat
std.residual <- (y - y.hat) / sd(residual)

H <- X %*% (solve(t(X) %*% X) %*% t(X))

DFITS <- std.residual * sqrt(diag(H) / (1 - diag(H)))
CookD <- (1 / p) * (diag(H) / (1 - diag(H))) * std.residual^2
```

```{r}
diagnosticos <- data.frame(Observacion = c(1:nrow(datos)),
                           Carbohidratos = y,
                           Ajustado = y.hat,
                           Residual = residual,
                           Residual.est = std.residual,
                           DFIT = DFITS,
                           CookD = CookD)
```

```{r QQ-Plot, fig.width=3, fig.height=3}
ggplot(data = diagnosticos, aes(sample = Residual.est)) +
  stat_qq(size = 3) +
  stat_qq_line() +
  xlab(label = "Cuantiles normales") +
  ylab(label = "Residuales estandarizados")
```

```{r Residuals, fig.width=3, fig.height=3}
ggplot(data = diagnosticos, aes(x = Ajustado, y = Residual.est)) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0) +
  xlab(label = "Carbohidratos ajustados") +
  ylab(label = "Residuales estandarizados")
```

```{r Cook`s Distance, fig.width=3, fig.height=3}
ggplot(data = diagnosticos, aes(x = Observacion, y = CookD)) +
  geom_point(size = 3) +
  xlab(label = "Órden de la observación") +
  ylab(label = "Distancia de Cook")
```

```{r DFITS, fig.width=3, fig.height=3}
ggplot(data = diagnosticos, aes(x = Observacion, y = DFIT)) +
  geom_point(size = 3) +
  xlab(label = "Órden de la observación") +
  ylab(label = "DFIT")
```

